---
title: "Basic Data Manipulation in R"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Subsetting Data
## Using Base R
```{r}
df <- iris
df
```

### Three ways of accessing values from a single column
```{r}
df$Sepal.Length
df[1]
df[,1]
```

### Rows
```{r}
df[1,]
```

### We can get more than just one row or column
```{r}
df[1:10,]
df[1:10, 1:2]
```

### We can do negative selection
```{r}
df[1:10, -1]
df[1:10, -(1:3)]
```

### We can add columns
```{r}
df$id <- 1:nrow(df)

df
```

### Let's try subsetting using a condition
```{r}
which(df$Species == "virginica")
```

### If we just use the numbers, it looks like this
```{r}
df[101:150,]
```

### Replace it with the `which()` function we used before
```{r}
df[which(df$Species == "virginica"), ]
```

### Save it as a new data frame
```{r}
test <- df[which(df$Species == "virginica"), ]

test
```

### Try it numerically (name, petal width, and ID of all which have a Sepal Length less than 6)
```{r}
test <- df[which(df$Sepal.Length < 6), 4:6]

test
```


## Using dplyr
```{r message=FALSE}
library(tidyverse)
```

### Three useful functions for subsetting
```{r eval = FALSE}
filter()
select()
mutate()
```

### A couple of other things from dplyr we'll talk about later
```{r eval = FALSE}
group_by()
summarize()
```

### Getting started with dplyr

#### `filter()` subsets observations (rows) based on a condition
```{r}
setosas <- filter(df, Species == "setosa")

setosas
```

#### `select()` subsets variables (columns)
```{r}
setosas_sepal <- select(setosas, Sepal.Length, Species, id)

setosas_sepal
```

### Combining those together with regular R syntax looks like this:
```{r}
setosas_sepal <- select(filter(df, Species == "setosa"), Sepal.Length, Species, id)

setosas_sepal
```

### Using the Pipe
```{r eval = FALSE}
# Example from Hadley Wickham
foo_foo <- little_bunny()

bop(scoop_up(hop_through(foo_foo, forest), field_mouse), head)

foo_foo %>%
  hop_through(forest) %>%
  scoop_up(field_mouse) %>%
  bop(head)
```

### We can apply this syntax to the iris data
```{r}
setosas <- df %>%
  filter(Species == "setosa")

setosas

setosas <- df %>%
  filter(Species == "setosa") %>%
  select(Sepal.Length, Species, id)

setosas

df2 <- iris %>%
  mutate(id = row_number())

df2
```

# Tidying data
### Tidying the iris dataset with `gather()`
```{r}
tidy <- df %>%
  gather(key = variable, value = value, Sepal.Length:Petal.Width)

tidy
```

### Trying it with something a little messier
```{r message=FALSE}
wide <- read_csv("http://dartgo.org/qbs_iris")
wide
```

### Using `gather()` and `separate()` to clean the data
```{r}
tidy <- wide %>%
  gather(key = bad_id, value = value, setosa_1:ncol(wide))

tidy

tidy <- wide %>%
  gather(key = bad_id, value = value, setosa_1:ncol(wide)) %>%
  separate(col = bad_id, into = c("species", "id"), sep = "_")

tidy
```

### Getting back to our original dataset
```{r}
df_normal <- wide %>%
  gather(key = bad_id, value = value, setosa_1:ncol(wide)) %>%
  separate(col = bad_id, into = c("species", "id"), sep = "_") %>%
  spread(key = variable, value = value)

df_normal
```

# Manipulating/Aggregating data
### Split - Apply - Combine </br>Using `group_by()` and `summarize()`
```{r}
species_means <- df %>%
  group_by(Species) %>%
  summarize(pl = mean(Petal.Length), pw = mean(Petal.Width), sl = mean(Sepal.Length), sw = mean(Sepal.Width))

species_means
```

### Using a Star Wars Example
```{r}
starwars
```

```{r}
sw_heights <- starwars %>%
  group_by(homeworld, species) %>%
  summarize(mean_height = mean(height))

sw_heights
```

